# Covariation

While the variation describes the behaviors within a variable, covariation describes the behavior between variables.

Covariation is the tendency for the values of two or more variables to vary together in a related way.

```{r echo=FALSE, include=FALSE}
library(tidyverse)
```


## Two categorical variables

### Contingency table

To characterize covaritation of two categorical variable, we can use contingency table to display the frequency.

```{r}
table(diamonds$cut, diamonds$color)

table(diamonds$cut, diamonds$color, diamonds$clarity)

xtabs(~ cut + color, diamonds)

xtabs(~ cut + color + clarity, diamonds)

ftable(xtabs(~ cut + color + clarity, diamonds))

diamonds %>%
  group_by(cut, color)%>%
  count( )%>%
  group_by(cut)%>%
  mutate(sum = sum(n))%>%
  mutate(proportion = n/sum,
         percentage = (n/sum)*100)
```

### Tile plot

You can visualize the frequency table with a tile plot.

```{r}

diamonds%>% 
  count(color, cut)%>%
  ggplot(aes(color, cut))+
  geom_tile(aes(fill=n))

ggplot(diamonds)+
  geom_count(aes(cut, color))


#install.packages("seriation")


```

## Categorical + continuous variable

The covariation of a categorical and a continuous variable can be visualized and explored by treating the categorical variable as the group factor. Then we can apply all the methods we learned when dealing with a continuous variable.

### Summary table
R offers a number of ways we can summarize the mean, sd of a continous variable as a function of one or more grouping variables.

```{r}
# solution 1
with(diamonds, tapply(price, cut, mean))

with(diamonds, tapply(price, list(cut, color, clarity), mean))


# solution 2
# install.packages("doBy")
library(doBy)
data = as.data.frame(diamonds)
head(summaryBy(price ~ cut + clarity + color , data = data, FUN = mean))

head(summaryBy(price  + carat ~ cut + clarity + color , data = data, FUN = mean))

head(summaryBy(price  + carat ~ cut +  color , data = data, FUN = c(mean, sd)))

# solution 3
diamonds%>%
  group_by(cut, clarity, color)%>%
  summarise(mean = mean(price),
            sd = sd(price))


```

### Central tendency (mean): Bar plots

```{r}
#bar plot
diamonds%>%
  group_by(cut)%>%
  summarise(mean = mean(price))%>%
  ggplot(aes(cut, mean))+
  geom_bar(stat="identity")

diamonds%>%
  group_by(cut)%>%
  summarise(mean = mean(price))%>%
  ggplot(aes(cut, mean, fill = cut))+
  geom_bar(stat="identity")

# xlim(4, 10) + ylim(4, 10)

diamonds%>%
  group_by(cut)%>%
  summarise(mean = mean(price),
            sd = sd(price))%>%
  ggplot(aes(cut, mean))+
  geom_bar(stat="identity")+
  geom_errorbar(aes(ymin = mean - sd,
                    ymax = mean + sd),
                width = .2, size = 0.7, position = position_dodge(.9))

```
### Spread: boxplot

```{r}
# boxplot

ggplot(diamonds, aes(cut, price))+
  geom_boxplot()



```


### Distribution: density plot 

We can color-code the density plot to represent the group factor. 




```{r}
library(tidyverse)

ggplot(diamonds, aes(price))+
  geom_freqpoly(binwidth = 500)

ggplot(diamonds, aes(price))+
  geom_freqpoly(aes(color = cut), binwidth = 500)

# standardized count where the area under each frequency polygon is one

ggplot(diamonds, aes(x = price, y =  ..density..))+
  geom_freqpoly(aes(color = cut), binwidth = 500)




```

model = model name

displ = engine displacement, in litres

year = year of manufacture

cyl = number of cylinders

trans = type of transmission

drv => f = front-wheel drive, r = rear wheel drive, 4 = 4wd

cty = city miles per gallon

hwy = highway miles per gallon

fl = fuel type

class = "type" of car


```{r}
summary(mpg)

ggplot(mpg)+
  geom_boxplot(aes(x = reorder(class, hwy, FUN = median), y = hwy))

ggplot(mpg)+
  geom_boxplot(aes(x = reorder(class, hwy, FUN = median), y = hwy))+
  coord_flip()

```


## Two continuous variables

### Scatter plots

The most common way we visualize two continuous variables is by using a scatter plot.

```{r}
ggplot(diamonds)+
  geom_point(aes(carat, price))

# add transparency
ggplot(diamonds)+
  geom_point(aes(carat, price), alpha = 1/100)

# bin two variables
ggplot(diamonds)+
  geom_bin2d(aes(carat, price))

#install.packages("hexbin")
ggplot(diamonds)+
  geom_hex(aes(carat, price))

#bin one variable
ggplot(diamonds,aes(carat, price))+
  geom_boxplot(aes(group = cut_width(carat, 0.1)))

ggplot(diamonds,aes(carat, price))+
  geom_boxplot(aes(group = cut_width(carat, 0.5)))


```
### Bin one or both continuous variables

Sometime we can bin one or both continuous variables to convert them into categorical variable(s). In those cases, we apply what we learn in dealing with categorical variables, such as tile plots or boxplots.

```{r}


#bin one variable
ggplot(diamonds,aes(carat, price))+
  geom_boxplot(aes(group = cut_width(carat, 0.1)))

ggplot(diamonds,aes(carat, price))+
  geom_boxplot(aes(group = cut_width(carat, 0.5)))

# bin two variables
ggplot(diamonds)+
  geom_bin2d(aes(carat, price))

#install.packages("hexbin")
ggplot(diamonds)+
  geom_hex(aes(carat, price))

```



